---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Preparing the environment for futher R Markdown file processing:
```{r setoptions, echo=TRUE, warning=FALSE}
require(knitr)
opts_chunk$set(echo = TRUE, cache = TRUE, cache.path = "cache/", fig.path = "figure/")
```

First, load necessary packages
```{r load_packages, echo=TRUE, warning=FALSE}
library(data.table)
library(reshape2)
library(plyr)
library(lattice)
```

Read the data into data frame
```{r readdata, echo=TRUE}
df <- read.csv('activity.csv')
```


## What is mean total number of steps taken per day?

Calculate the total number of steps taken per day
```{r totalsteps, echo=TRUE}
nsday <- ddply(df, "date", summarise,
               N.steps = sum(steps, na.rm=TRUE) 
               )
ndays <- dim(nsday)[1]
print(head(nsday,5))
print(tail(nsday,5))
```

Make a histogram of the total number of steps taken each day
```{r histogram, fig.height=5, fig.width=6, echo=TRUE}
par(mar = c(5, 4, 1, 1), las = 1)
hist(nsday$N.steps,
     breaks = 10,
     density = 50,
     angle = 45,
     main = "Histogram of number of steps per day",
     xlim = c(0, 25000),
     ylim = c(0, 20),
     xlab = "Number of steps per day",
     labels = TRUE
     )
```

Calculate and report the mean and median of the total number of steps taken per day
```{r mean&median, echo=TRUE}
nsday.mean <- mean(nsday$N.steps)
nsday.median <- median(nsday$N.steps)
```

The **mean** of the total number of steps taken per day is **`r round(nsday.mean)`**

The **median** of the total number of steps taken per day is **`r nsday.median`**


## What is the average daily activity pattern?

Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

First, compute the average number of steps taken at each 5-min interval
```{r aversteps, echo=TRUE}
nsint <- ddply(df, "interval", summarise,
               N.steps = mean(steps, na.rm = TRUE)
               )
nint <- dim(nsint)[1]
print(head(nsint,5))
print(tail(nsint,5))
```

Plot the average number of steps taken against all intervals
```{r plotall, fig.height=4, fig.width=8, echo=TRUE}
par(mar = c(5, 4, 1, 1), las = 1)
plot(nsint$interval, nsint$N.steps,
     type = "l",
     main = "Number of steps as a function of interval",
     xlim = c(0, 2400),
     xlab = "Interval",
     ylab = "Mean number of steps"
     )
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r maxsteps, echo=TRUE}
stepmax <- max(nsint$N.steps)
imax <- nsint$N.steps==stepmax
intmax <- nsint$interval[imax]
```

The interval **`r intmax`**  contains the maximum number of steps (**`r round(stepmax)`**)

## Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r numberNAs, echo=TRUE}
N.NAs.s <- sum(as.numeric(is.na(df$steps)))
N.NAs.i <- sum(as.numeric(is.na(df$interval)))
N.NAs.d <- sum(as.numeric(is.na(df$date)))
print(c(N.NAs.s, N.NAs.i, N.NAs.d))
```

The total number of missing values in the dataset is **`r N.NAs.s`** that comprise **`r N.NAs.s/nint`** days

Let's use the mean for each 5-minute interva to fill the gaps

Create a new dataset that is equal to the original dataset but with the missing data filled in
```{r replaceNAs, echo=TRUE}
df.mod <- df
nsint.long <- rep(nsint$N.steps, ndays)
df.mod$steps[is.na(df$steps)] <- nsint.long[is.na(df$steps)]
print(summary(df.mod))
print(summary(df))
```

Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

Calculate the total number of steps taken per day
```{r totalsteps.mod, echo=TRUE}
nsday.mod <- ddply(df.mod, "date", summarise,
               N.steps = sum(steps, na.rm=TRUE) 
               )
print(head(nsday.mod,5))
print(tail(nsday.mod,5))
```

Make a histogram of the total number of steps taken each day
```{r histogram.mod, fig.height=5, fig.width=6, echo=TRUE}
par(mar = c(5, 4, 1, 1), las = 1)
hist(nsday.mod$N.steps,
     breaks = 10,
     density = 50,
     angle = 45,
     main = "Histogram of number of steps per day with NA imputed",
     xlim = c(0, 25000),
     ylim = c(0, 25),
     xlab = "Number of steps per day",
     labels = TRUE
     )
```

Calculate and report the mean and median of the total number of steps taken per day
```{r mean&median.mod, echo=TRUE}
nsday.mea <- mean(nsday.mod$N.steps)
nsday.med <- median(nsday.mod$N.steps)
```

The **mean** of the total number of steps taken per day is **`r as.integer(nsday.mea)`**

The **median** of the total number of steps taken per day is **`r as.integer(nsday.med)`**


## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r parseweekdays, echo=TRUE}
weekday7 <- weekdays(as.Date(df.mod$date))
weekday2 <- rep("weekday", 17568)
weekday2[weekday7 == "Saturday" | weekday7 == "Sunday"] <- "weekend"
#weekday2 <- rep("weekend", 17568)
#weekday2[weekday7 != "Saturday" | weekday7 != "Sunday"] <- "weekday"
df.mod$weekday <- weekday2
df.mod$stepswkd <- rep(NaN, 17568)
df.mod$stepswkd[df.mod$weekday=="weekday"] <- df.mod$steps[df.mod$weekday=="weekday"]
df.mod$stepswnd <- rep(NaN, 17568)
df.mod$stepswnd[df.mod$weekday=="weekend"] <- df.mod$steps[df.mod$weekday=="weekend"]
```

Compute the average number of steps taken at each 5-min interval during weekdays and weekends
```{r averstepswkd, echo=TRUE}
nsintwd <- ddply(df.mod, "interval", summarise,
                 weekday = mean(stepswkd, na.rm=TRUE),
                 weekend = mean(stepswnd, na.rm=TRUE)
                 )
print(head(nsintwd,5))
print(tail(nsintwd,5))
```

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)
```{r plotweeknd, echo=TRUE}
nsintwd.molten <- melt(nsintwd, id.vars=c("interval"))
xyplot(
      nsintwd.molten$value ~ nsintwd.molten$interval | 
      nsintwd.molten$variable, panel =           
      function(x, y, ...){panel.xyplot(x, y, ...)},
      type="l",
      xlab="Intervals",
      ylab="Number of steps",
      layout=c(1,2)
      )
```

